---
title: "Practical: Data Visualisation"
---

Create effective visualizations to explore and communicate your data insights.

## Exercise: Visualizing Data

### Visualizing in R

Create a simple plot to understand the relationship between variables:

```{r}
#| eval: false

library(ggplot2)

# Scatter plot: mpg vs weight
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Fuel Efficiency vs Car Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()
```

**Your Task**: 
- Create the plot above
- What pattern do you observe?
- Try changing the x variable to `hp` (horsepower)

### Visualizing in Python

```{python}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

# Load data
df = sns.load_dataset('mpg')

# Create scatter plot
plt.figure(figsize=(10, 6))
sns.scatterplot(data=df, x='weight', y='mpg')
plt.title('Fuel Efficiency vs Car Weight')
plt.xlabel('Weight')
plt.ylabel('Miles per Gallon')
plt.show()
```

## Bar Plots by Category

### In R

```{r}
#| eval: false

# Create a bar plot of trips by mode
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Cars by Cylinder Count",
       x = "Cylinders",
       y = "Number of Cars") +
  theme_minimal()
```

### In Python

```{python}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

df = sns.load_dataset('mpg')

# Count by origin
sns.countplot(data=df, x='origin')
plt.title('Car Count by Origin')
plt.show()
```

## Download the Code

You can download the code for this exercise:

::: {.callout-tip}
## Download Options
- [Download R code](../../code/practical-data-visualisation.R)
- [Download Python code](../../code/practical-data-visualisation.py)
:::

## Common Issues & Solutions

::: {.callout-note}
## Plot not displaying?
Make sure you're in an interactive environment. In scripts, use `plt.show()` in Python or print the ggplot object in R.
:::

## Key Takeaways

After completing this exercise, you should be able to:

✅ Create scatter plots and trend lines  
✅ Create bar plots and histograms  
✅ Customize plot titles and labels  
✅ Interpret patterns in visualizations
