---
title: Practical Exercises
---

Let's get hands-on with data science! These exercises will give you practical experience with core data science tasks.

## Exercise 1: Loading and Exploring Data

### In R

Let's work with a built-in dataset about cars:

```{r}
#| eval: false

# Load tidyverse
library(tidyverse)

# Look at the mtcars dataset
head(mtcars)

# Get basic statistics
summary(mtcars)

# Check the structure
str(mtcars)
```

**Your Task**: Explore the dataset and answer:
- How many cars are in the dataset?
- What variables are included?
- What is the average miles per gallon (mpg)?

### In Python

```{python}
#| eval: false

import pandas as pd
import seaborn as sns

# Load a sample dataset
df = sns.load_dataset('mpg')

# Explore the data
print(df.head())
print(df.describe())
print(df.info())
```

## Exercise 2: Data Visualization

### Visualizing in R

Create a simple plot to understand the relationship between variables:

```{r}
#| eval: false

library(ggplot2)

# Scatter plot: mpg vs weight
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Fuel Efficiency vs Car Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()
```

**Your Task**: 
- Create the plot above
- What pattern do you observe?
- Try changing the x variable to `hp` (horsepower)

### Visualizing in Python

```{python}
#| eval: false

import matplotlib.pyplot as plt
import seaborn as sns

# Load data
df = sns.load_dataset('mpg')

# Create scatter plot
plt.figure(figsize=(10, 6))
sns.scatterplot(data=df, x='weight', y='mpg')
plt.title('Fuel Efficiency vs Car Weight')
plt.xlabel('Weight')
plt.ylabel('Miles per Gallon')
plt.show()
```

## Exercise 3: Data Manipulation

### Filtering and Summarizing in R

```{r}
#| eval: false

library(tidyverse)

# Filter cars with good fuel efficiency (mpg > 20)
efficient_cars <- mtcars %>%
  filter(mpg > 20)

# Calculate average mpg by cylinder count
mpg_by_cyl <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    avg_mpg = mean(mpg),
    count = n()
  )

print(mpg_by_cyl)
```

**Your Task**:
- How many cars have mpg > 20?
- Which cylinder count has the best average fuel efficiency?
- Filter for cars with automatic transmission (`am == 0`)

### Filtering in Python

```{python}
#| eval: false

import pandas as pd
import seaborn as sns

df = sns.load_dataset('mpg')

# Filter cars from the USA
usa_cars = df[df['origin'] == 'usa']

# Group by year and calculate average mpg
mpg_by_year = df.groupby('model_year')['mpg'].mean()

print(mpg_by_year)
```

## Exercise 4: Working with Transport Data

### Creating Simple Transport Analysis in R

```{r}
#| eval: false

library(tidyverse)

# Create a simple trip dataset
trips <- data.frame(
  trip_id = 1:10,
  distance_km = c(5.2, 3.1, 7.8, 2.5, 15.3, 4.2, 6.7, 8.9, 3.4, 5.8),
  mode = c("bus", "walk", "car", "walk", "train", "bus", "car", "bike", "walk", "bus"),
  duration_min = c(25, 15, 20, 10, 45, 20, 18, 30, 12, 22)
)

# Calculate average speed by mode
trip_summary <- trips %>%
  mutate(speed_kmh = (distance_km / duration_min) * 60) %>%
  group_by(mode) %>%
  summarise(
    avg_speed = mean(speed_kmh),
    avg_distance = mean(distance_km),
    count = n()
  )

print(trip_summary)
```

**Your Task**:
- Which transport mode has the highest average speed?
- Calculate the total distance traveled across all trips
- Create a bar plot showing trip counts by mode

### Visualization

```{r}
#| eval: false

# Create a bar plot of trips by mode
ggplot(trips, aes(x = mode)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Trips by Transport Mode",
       x = "Transport Mode",
       y = "Number of Trips") +
  theme_minimal()
```

## Bonus Challenge

If you finish early, try this:

1. **Find a real dataset** online (e.g., from [data.gov.uk](https://www.data.gov.uk/) or [Our World in Data](https://ourworldindata.org/))
2. **Load it** into R or Python
3. **Create 3 visualizations** that tell a story about the data
4. **Share your findings** with the group!

## Common Issues & Solutions

::: {.callout-note}
## Error: Package not found
Make sure you've installed the package first:
```r
install.packages("package_name")
```
:::

::: {.callout-note}
## Error: Object not found
Check that you've run all previous code chunks in order. Variables from earlier chunks need to be created first!
:::

::: {.callout-tip}
## Getting Stuck?
- Check the documentation: `?function_name` in R
- Use AI tools like Copilot or ChatGPT
- Ask your neighbor or the instructor!
:::

## Key Takeaways

After completing these exercises, you should be able to:

✅ Load and explore datasets  
✅ Create basic visualizations  
✅ Filter and summarize data  
✅ Calculate derived variables  
✅ Understand transport data structures  

## Resources for Practice

- [R for Data Science (free book)](https://r4ds.hadley.nz/)
- [Python Data Science Handbook](https://jakevdp.github.io/PythonDataScienceHandbook/)
- [TidyTuesday](https://github.com/rfordatascience/tidytuesday) - Weekly data challenges
