---
title: Prerequisites
---

Before starting this practical session, you'll need to set up a few essential tools. Don't worry—we'll keep this brief!

## Required Software

### 1. R and RStudio

**R** is a programming language widely used for statistical computing and data science.

- Download R from [CRAN](https://cran.r-project.org/)
- Download RStudio Desktop (free) from [Posit](https://posit.co/download/rstudio-desktop/)

### 2. Python (Optional but Recommended)

**Python** is another popular language for data science.

- Download Python from [python.org](https://www.python.org/downloads/) (version 3.8 or higher)
- Or install via [Anaconda](https://www.anaconda.com/download) (includes many data science packages)

### 3. Git and GitHub Account

**Git** is version control software, and **GitHub** is a platform for hosting code.

- Download Git from [git-scm.com](https://git-scm.com/)
- Create a free GitHub account at [github.com](https://github.com/)

## Essential R Packages

Install these R packages by running in RStudio:

```{r}
#| eval: false

# Data manipulation and visualization
install.packages("tidyverse")

# For working with transport data
install.packages("sf")
```

## Essential Python Packages (Optional)

If using Python, install these packages:

```{bash}
#| eval: false

pip install pandas numpy matplotlib seaborn
```

## Checking Your Setup

### Test R Installation

Open RStudio and run:

```{r}
#| eval: false

# Check R version
R.version.string

# Test tidyverse
library(tidyverse)
```

### Test Python Installation (Optional)

Open a terminal/command prompt and run:

```{bash}
#| eval: false

python --version
pip list
```

## Getting Help

If you encounter installation issues:

- Check the [RStudio Support](https://support.posit.co/hc/en-us) page
- Visit [Stack Overflow](https://stackoverflow.com/) for troubleshooting
- Ask during the session!

::: {.callout-tip}
## Pro Tip
Don't spend too much time troubleshooting installation issues before the session. We can help you during the practical!
:::
