--- 
title: "Statistics - Worked Solutions"
format:
  html:
    code-fold: true
---

This page presents the worked solutions for the statistics tasks, showing how to perform the same analyses in both R and Python.

::: callout-warning
## Before you proceed...

These are worked solutions. We strongly recommend attempting to debug the R and Python problem scripts in the [Tasks section](index.qmd) before reviewing these solutions. Learning to debug is a crucial data science skill!
:::

::: {.callout-note}
## High-Performance Alternatives: Polars and DuckDB

The Python solutions below use **pandas** for seamless integration with `seaborn` and `statsmodels`. When working with larger-than-RAM datasets or complex data engineering pipelines, modern alternatives to explore include:

-   **[Polars](https://docs.pola.rs/)**: A fast, multi-threaded DataFrame library built in Rust.
-   **[DuckDB](https://duckdb.org/)**: An in-process SQL analytical engine for querying files directly without database overhead.
:::

## Task 1: Excel Data Analysis

This task involves reading data from an Excel file (`ParkRunPerformanceData.xlsx`), cleaning it, calculating summaries, and visualizing the distribution of run times.

::: panel-tabset

### R

```{r}
#| label: excel-task-r
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
#| fig-cap: "Various Outputs from Excel Task (R)"

# Loading the necessary libraries
library(tidyverse)
library(readxl)

# Assigning the path to a variable
path_to_file <- "../00_data/ParkRunPerformanceData.xlsx" # Corrected path

# Reading data from an Excel file
data <- read_excel(path = path_to_file,
                   sheet = "Sheet1",
                   col_types = c("date", "numeric"))

# Exploring the data
cat("---\n--- Head of Data ---\n---")
print(head(data))

cat("\n--- Structure of Data ---")
str(data)

cat("\n--- Summary Statistics ---")
summary(data)


# Rename the columns to ensure consistency and ease manipulation
names(data) <- c("date","runtime")

# Sorting the data
data_sorted <- data |> arrange(runtime)
cat("\n--- Sorted Data (Ascending) ---")
print(head(data_sorted))

data_sorted_inv <- data |> arrange(-runtime)
cat("\n--- Sorted Data (Descending) ---")
print(head(data_sorted_inv))


# Calculating the summaries manually 
summary_dates <- 
  data |>
  summarise(min_date = min(date),
            max_date = max(date))
cat("\n--- Summary Dates ---")
print(summary_dates)

summary_runtimes <- 
  data |>
  summarise(
            count = n(),
            mean_runtime = mean(runtime),
            slowest = max(runtime),
            fastest = min(runtime))
cat("\n--- Summary Runtimes ---")
print(summary_runtimes)

# Rounding the run times to the nearest minute
data_rounded <- data |> 
  mutate(runtime_mins = round(x = runtime, digits = 0))
cat("\n--- Data with Rounded Runtimes ---")
print(head(data_rounded))


# Counting the frequencies for each value
freq <- data_rounded |> 
  count(runtime_mins)
cat("\n--- Frequencies ---")
print(freq)

# A quick histogram
hist(data_rounded$runtime_mins,breaks = 14:33, main = "Quick Histogram of Run Times")

# A nicer histogram
data |> 
  ggplot(aes(x = runtime))+
  geom_histogram(binwidth = 1, col = "white",fill = "steelblue", alpha = 0.7)+
  labs (x = "Run time in seconds",
        y = "frequency",
        title = "Park Run Times Distribution",
        subtitle = "Records from Aug 2012 to Aug 2015",
        caption = "Source: Andrew Tomlinson")+
  scale_x_continuous(breaks = 14:33)+
  geom_hline(yintercept = 0,linewidth = 1,col = "grey30")+
  theme_minimal()+
  theme(title = element_text(size = 15))

# Have the run times improved?
  # A different exploration
data |> 
  ggplot(aes(x = date, y = runtime))+
  geom_point() +  
  geom_smooth(method = "lm")
```

### Python

```{r}
#| include: false
# Install Python packages from requirements.txt
reticulate::virtualenv_install(
  packages = readLines("../requirements.txt"),
  ignore_installed = TRUE
)
```

```{python}
#| label: excel-task-py
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
#| fig-cap: "Various Outputs from Excel Task (Python)"

import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
import matplotlib.dates as mdates

sns.set_theme(style="whitegrid")

# Assigning the path to a variable
path_to_file = "../00_data/ParkRunPerformanceData.xlsx" # Corrected path

# Reading data from an Excel file
data = pd.read_excel(path_to_file, sheet_name="Sheet1")

# Exploring the data
print("---\n--- Head of Data ---\n---")
print(data.head())

print("\n--- Info ---")
data.info()

print("\n--- Summary Statistics ---")
print(data.describe())

# Rename the columns
if len(data.columns) >= 2:
    data.columns = ["date", "runtime"]

# Sorting the data
data_sorted = data.sort_values(by="runtime")
print("\n--- Sorted Data (Ascending) ---")
print(data_sorted.head())

data_sorted_inv = data.sort_values(by="runtime", ascending=False)
print("\n--- Sorted Data (Descending) ---")
print(data_sorted_inv.head())

# Calculating summaries
summary_dates = pd.Series({
    "min_date": data["date"].min(),
    "max_date": data["date"].max()
})
print("\n--- Summary Dates ---")
print(summary_dates)

summary_runtimes = pd.Series({
    "count": data["runtime"].count(),
    "mean_runtime": data["runtime"].mean(),
    "slowest": data["runtime"].max(),
    "fastest": data["runtime"].min()
})
print("\n--- Summary Runtimes ---")
print(summary_runtimes)

# Rounding the run times
data["runtime_mins"] = data["runtime"].round(0)
print("\n--- Data with Rounded Runtimes ---")
print(data.head())

# Frequencies
freq = data["runtime_mins"].value_counts().sort_index()
print("\n--- Frequencies ---")
print(freq)

# Nicer Histogram
plt.figure(figsize=(10, 6))
sns.histplot(
    data=data, 
    x="runtime", 
    binwidth=1, 
    color="steelblue", 
    edgecolor="white", 
    alpha=0.7
)
plt.title("Park Run Times Distribution\nRecords from Aug 2012 to Aug 2015", fontsize=15)
plt.xlabel("Run time in seconds")
plt.ylabel("frequency")
plt.xlim(14, 33)
plt.axhline(0, color="grey", linewidth=1)
plt.figtext(0.8, 0.01, "Source: Andrew Tomlinson", wrap=True, horizontalalignment='center', fontsize=10)
plt.show()

# Have the run times improved?
plt.figure(figsize=(10, 6))
data_reg = data.copy()
data_reg["date_num"] = mdates.date2num(data_reg["date"])

sns.regplot(
    data=data_reg, 
    x='date_num', 
    y="runtime", 
    scatter_kws={'s':10}, 
    line_kws={'color':'blue'}
)
# Fix x-axis to show dates
ax = plt.gca()
ax.xaxis.set_major_formatter(mdates.DateFormatter('%Y-%m'))
plt.title("Run Times over Date")
plt.show()
```

:::

## Task 2: SPSS Data Analysis

This task analyzes running data originally handled in SPSS (`RunningData.xlsx`), including filtering, grouping, and statistical tests.

::: panel-tabset

### R

```{r}
#| label: spss-task-r
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
#| fig-cap: "Various Outputs from SPSS Task (R)"

library(tidyverse)
library(readxl)

path_to_file <- "../00_data/RunningData.xlsx"
data <- read_excel(path = path_to_file, sheet = "Sheet1")

# Rename columns
names(data) <- c("position", "time", "age_cat", "gender", "prev_runs")

# Summaries
cat("---\n--- Summary of Time ---")
summary(data$time)
cat("\n--- Unique Age Categories ---")
unique(data$age_cat)

# Subset only adults 
children_cats <- c("10","11-14","15-17")
data_adults <- data |> 
  filter(!age_cat %in% children_cats)

cat("\n--- Unique Adult Age Categories ---")
unique(data_adults$age_cat)
cat("\n--- Summary of Adult Time ---")
summary(data_adults$time)

# Analysis by gender
cat("\n--- Analysis by Gender ---")
data_adults |> 
  group_by(gender) |> 
  summarise(min = min(time),
            mean = mean(time),
            median = median(time),
            max = max(time))

# Comparing distributions
data_adults |> 
  ggplot(aes(x = time, fill = gender))+
  geom_histogram(col = "white")+
  facet_grid(gender~.)

data_adults |> 
  ggplot(aes(x = time, col = gender))+
  geom_density()

data_adults |> 
  ggplot(aes(x = time, col = gender))+
  geom_boxplot()

data_adults |> 
  ggplot(aes(x = time, y = gender, col = gender))+
  geom_violin()

## Statistical tests
# Extracting the data
times_female_adults <- data_adults |> filter(gender == "F") |> pull(time)
times_male_adults <- data_adults |> filter(gender == "M") |> pull(time)

cat("\n--- T-Test Results (Male vs Female) ---")
t.test(times_male_adults, times_female_adults)

# Previous runs vs times
data_adults |> 
  ggplot(aes(x = prev_runs, y = time))+
  geom_point()+
  geom_smooth(method = "lm")

data_adults |> 
  ggplot(aes(x = prev_runs, y =time, col = gender))+
  geom_point() +
  geom_smooth(method = "lm")

cat("\n--- Correlation Test (Time vs Previous Runs) ---")
cor.test(data_adults$time, data_adults$prev_runs)

# Finding the median of prev runs
median_prev_runs <- median(data_adults$prev_runs)
cat("\n--- Median Previous Runs ---")
print(median_prev_runs)

data_adults_pr_gr <- data_adults |> 
  mutate(pr_gr = prev_runs>=median_prev_runs)

cat("\n--- Data with pr_gr group ---")
print(head(data_adults_pr_gr))

# a quick visual check
data_adults_pr_gr |> 
  ggplot(aes(prev_runs,fill = pr_gr))+
  geom_histogram()

# Comparing times
data_adults_pr_gr |> 
  ggplot(aes(x = time, col = pr_gr))+
  geom_boxplot()

# Linear Model
# Extract first two digits of age category for numeric age proxy
data_adults_lm <- data_adults |> 
  mutate(age = str_extract(age_cat, '^\\d{2}') |> as.numeric())  

my_linear_model <- lm(formula = "time ~ age + gender + prev_runs", data = data_adults_lm) 
cat("\n--- Linear Model Summary ---")
summary(my_linear_model)
```

### Python

```{python}
#| label: spss-task-py
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
#| fig-cap: "Various Outputs from SPSS Task (Python)"

import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
import scipy.stats as stats
import statsmodels.formula.api as smf

sns.set_theme(style="whitegrid")

path_to_file = "../00_data/RunningData.xlsx"
data = pd.read_excel(path_to_file, sheet_name="Sheet1")

# Rename columns
if len(data.columns) >= 5:
    data.columns = ["position", "time", "age_cat", "gender", "prev_runs"]

print("---\n--- Head ---")
print(data.head())
print("\n--- Info ---")
data.info()
print("\n--- Summary ---")
print(data.describe())
print("\n--- Time Summary ---")
print(data["time"].describe())

# Subset adults
children_cats = ["10", "11-14", "15-17"]
data_adults = data[~data["age_cat"].isin(children_cats)].copy()

print("\n--- Adult Age Categories ---\n")
print(data_adults["age_cat"].unique())
print("\n--- Adult Time Summary ---")
print(data_adults["time"].describe())

# Analysis by gender
print("\n--- Analysis by Gender ---")
summary_gender = data_adults.groupby("gender")["time"].agg(["min", "mean", "median", "max"])
print(summary_gender)

# Comparing distributions
g = sns.FacetGrid(data_adults, row="gender", hue="gender", aspect=2, height=3)
g.map(sns.histplot, "time", edgecolor="white")
plt.show()

plt.figure(figsize=(10, 6))
sns.kdeplot(data=data_adults, x="time", hue="gender")
plt.title("Density by Gender")
plt.show()

plt.figure(figsize=(10, 6))
sns.boxplot(data=data_adults, x="time", hue="gender")
plt.title("Boxplot by Gender")
plt.show()

plt.figure(figsize=(10, 6))
sns.violinplot(data=data_adults, x="time", y="gender", hue="gender")
plt.title("Violin Plot by Gender")
plt.show()

## Statistical tests
# Extracting the data
times_female_adults = data_adults[data_adults["gender"] == "F"]["time"]
times_male_adults = data_adults[data_adults["gender"] == "M"]["time"]

print("\n--- Female Times (Head) ---")
print(times_female_adults.head())
print("\n--- Male Times (Head) ---")
print(times_male_adults.head())

t_stat, p_val = stats.ttest_ind(times_male_adults, times_female_adults)
print(f"\n--- T-Test Results ---\nt-statistic: {t_stat}\np-value: {p_val}")

# Previous runs vs times
plt.figure(figsize=(10, 6))
sns.regplot(data=data_adults, x="prev_runs", y="time")
plt.title("Time vs Previous Runs")
plt.show()

plt.figure(figsize=(10, 6))
sns.lmplot(data=data_adults, x="prev_runs", y="time", hue="gender")
plt.title("Time vs Previous Runs by Gender")
plt.show()

# Correlation
corr, p_corr = stats.pearsonr(data_adults["time"], data_adults["prev_runs"])
print(f"\n--- Correlation Test ---\ncorrelation: {corr}\np-value: {p_corr}")

# Finding the median of prev runs
median_prev_runs = data_adults["prev_runs"].median()
print(f"\nMedian Previous Runs: {median_prev_runs}")

data_adults["pr_gr"] = data_adults["prev_runs"] >= median_prev_runs

print("\n--- Data with pr_gr group ---")
print(data_adults[["prev_runs", "pr_gr"]].head())

# a quick visual check
plt.figure(figsize=(10, 6))
sns.histplot(data=data_adults, x="prev_runs", hue="pr_gr", multiple="stack")
plt.title("Previous Runs Split")
plt.show()

# Comparing times
plt.figure(figsize=(10, 6))
sns.boxplot(data=data_adults, x="time", hue="pr_gr")
plt.title("Time by Previous Runs Group")
plt.show()

# Linear Model
data_adults["age"] = data_adults["age_cat"].str.extract(r"^(\d{2})").astype(float)

model = smf.ols(formula="time ~ age + gender + prev_runs", data=data_adults)
results = model.fit()
print("\n--- Linear Model Summary ---\n")
print(results.summary())
```
:::
