---
format: revealjs
---

# Data Visualisation

Tell stories with your data

------------------------------------------------------------------------

## Why Visualise?

- Understand patterns quickly
- Communicate findings effectively
- Identify outliers and anomalies
- Explore relationships

------------------------------------------------------------------------

## Scatter Plots

Show relationships between two continuous variables

------------------------------------------------------------------------

### Step 1: Create Basic Figure

Start with the data and create a figure:

```python
import matplotlib.pyplot as plt
import seaborn as sns
import pandas as pd

# Read data
df = sns.load_dataset('mpg')

# Step 1: Create figure and axis
fig, ax = plt.subplots(figsize=(10, 6))
```

------------------------------------------------------------------------

### Step 2: Add Points

Add points to visualise individual observations:

```python
# Step 2: Create scatter plot with custom styling
ax.scatter(df['weight'], df['mpg'], s=100, alpha=0.6, color='steelblue')
```

------------------------------------------------------------------------

### Step 3: Add Labels and Theme

Add titles, axis labels, and a grid to make the plot publication-ready:

```python
# Add labels and theme
ax.set_xlabel('Weight')
ax.set_ylabel('Miles per Gallon')
ax.set_title('Fuel Efficiency vs Weight')
ax.grid(alpha=0.3)
```

------------------------------------------------------------------------

### Step 4: Add Trend Line

Add a regression line to show the pattern:

```python
# Add regression line
sns.regplot(data=df, x='weight', y='mpg', ax=ax, 
            scatter=False, color='red', line_kws={'label':'Trend'})
ax.legend()
```

------------------------------------------------------------------------

### Step 5: Final Plot

View the complete publication-ready plot:

```python
plt.tight_layout()
plt.show()
```

------------------------------------------------------------------------

## Bar Plots

Show counts or comparisons across categories

------------------------------------------------------------------------

### Step 1: Basic Bar Plot

Create bars for each category:

```python
import matplotlib.pyplot as plt
import seaborn as sns

df = sns.load_dataset('mpg')

# Create figure
fig, ax = plt.subplots(figsize=(10, 6))

# Create bar plot
df['cylinders'].value_counts().sort_index().plot(
    kind='bar', ax=ax, color='steelblue', alpha=0.7
)
```

------------------------------------------------------------------------

### Step 2: Add Labels

Add titles and axis labels:

```python
# Add labels
ax.set_title('Number of Cars by Cylinders')
ax.set_xlabel('Cylinders')
ax.set_ylabel('Count')
plt.show()
```

------------------------------------------------------------------------

## Box Plots

Show distributions and identify outliers

------------------------------------------------------------------------

### Step 1: Basic Box Plot

Create a box plot to show distribution:

```python
import matplotlib.pyplot as plt
import seaborn as sns

df = sns.load_dataset('mpg')

# Create figure
fig, ax = plt.subplots(figsize=(10, 6))

# Create box plot
sns.boxplot(data=df, x='cylinders', y='mpg', ax=ax, palette='Set2')
```

------------------------------------------------------------------------

### Step 2: Customise Box Plot

Add individual points and improve aesthetics:

```python
# Add individual points
sns.stripplot(data=df, x='cylinders', y='mpg', ax=ax, 
              color='black', alpha=0.3, size=3)

# Add labels
ax.set_title('MPG Distribution by Number of Cylinders')
ax.set_xlabel('Cylinders')
ax.set_ylabel('Miles per Gallon')
plt.show()
```

------------------------------------------------------------------------

## Using Plots

Plots can also be assigned to objects for further modification or saving:

```python
# Save plot to file
fig.savefig('my_boxplot.png', dpi=300, bbox_inches='tight')
```

------------------------------------------------------------------------

## Exporting Plots
You can save your plots using `fig.savefig()`:

```python
# Save the final plot as a PNG file
fig.savefig("final_plot.png", dpi=300, bbox_inches='tight')
```
---

## Key Resources

- [Python Data Science Handbook - Visualization](https://jakevdp.github.io/PythonDataScienceHandbook/04.00-introduction-to-matplotlib.html)
- [Matplotlib Documentation](https://matplotlib.org/)
- [Seaborn Documentation](https://seaborn.pydata.org/)
- [Visualization for Social Data Science](https://vis4sds.github.io/vis4sds/)

------------------------------------------------------------------------

## Key Takeaways

✅ Start with a basic figure and build incrementally

✅ Choose the right plot type for your data

✅ Customise colours, size, and transparency strategically

✅ Add informative titles and axis labels

✅ Use themes/styles to improve presentation

✅ Layer multiple elements for richer visualisations

------------------------------------------------------------------------

## Modern Python Visualisation

Beyond standard Matplotlib and Seaborn:

-   **[Seaborn Objects API](https://seaborn.pydata.org/tutorial/objects_overview.html):** A grammar-of-graphics interface (`import seaborn.objects as so`) that closely mirrors `ggplot2`.
-   **[Plotly](https://plotly.com/python/):** Interactive web figures with zoom, pan, and hover tooltips for Quarto reports.

------------------------------------------------------------------------

## R Content (Optional)

If you're interested in R, here are equivalent concepts

------------------------------------------------------------------------

### Scatter Plot in R

Step-by-step construction of a scatter plot:

```{r}
#| echo: true
library(ggplot2)

# Create the basic ggplot object with axes
ggplot(mtcars, aes(x = wt, y = mpg))
```

------------------------------------------------------------------------

### Step 2: Add Points

Add points to visualise individual observations:

```{r}
#| echo: true
# Add points to show the relationship
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

------------------------------------------------------------------------

### Step 3: Customise Points

Adjust size, transparency, and colour:

```{r}
#| echo: true
# Make points larger and semi-transparent
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, alpha = 0.6, colour = "steelblue")
```

------------------------------------------------------------------------

### Step 4: Add Trend Line

Add a smoothing layer to show the pattern:

```{r}
#| echo: true
# Add linear regression line
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, alpha = 0.6, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "red")
```

------------------------------------------------------------------------

### Step 5: Add Labels and Theme

Make it publication-ready with titles and formatting:

```{r}
#| echo: true
# Build the complete scatter plot
scatter_plot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, alpha = 0.6, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  labs(
    title = "Fuel Efficiency vs Weight",
    subtitle = "mtcars dataset",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    caption = "Data: Henderson & Velleman (1981)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

------------------------------------------------------------------------

### Final Scatter Plot

View the complete publication-ready plot:

```{r}
#| echo: false
scatter_plot
```

------------------------------------------------------------------------

## Bar Plots in R

Show counts or comparisons across categories

------------------------------------------------------------------------

### Step 1: Basic Bar Plot

Create bars for each category:

```{r}
#| echo: true
# Create basic bar plot
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()
```

------------------------------------------------------------------------

### Step 2: Add Colour

Customise the appearance with colours:

```{r}
#| echo: true
# Add custom fill colour
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue", colour = "black", alpha = 0.7)
```

------------------------------------------------------------------------

### Step 3: Add Labels

Add titles and axis labels:

```{r}
#| echo: true
# Build the complete bar plot
bar_plot <- ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue", colour = "black", alpha = 0.7) +
  labs(
    title = "Number of Cars by Cylinders",
    x = "Cylinders",
    y = "Count"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

### Final Bar Plot

View the complete publication-ready plot:

```{r}
#| echo: false
bar_plot
```

------------------------------------------------------------------------

## Box Plots in R

Show distributions and identify outliers

------------------------------------------------------------------------

### Step 1: Basic Box Plot

Create a box plot to show distribution:

```{r}
#| echo: true
# Create basic box plot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()
```

------------------------------------------------------------------------

### Step 2: Customise Box Plot

Add colours and improve aesthetics:

```{r}
#| echo: true
# Customised box plot with colour
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.3)  # Add individual points
```

------------------------------------------------------------------------

### Step 3: Add Labels

Make it clear and publication-ready:

```{r}
#| echo: true
# Complete box plot with labels
my_boxplot <- ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "MPG Distribution by Number of Cylinders",
    x = "Cylinders",
    y = "Miles per Gallon",
    fill = "Cylinders"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

### Final Box Plot

View the complete publication-ready plot:

```{r}
#| echo: false
my_boxplot
```

------------------------------------------------------------------------

## Using Plots in R

Plots can also be assigned to objects for further modification or saving:

```{r}
#| echo: true
# Save plot to object
final_plot <- my_boxplot + theme_classic()

final_plot
```

------------------------------------------------------------------------

## Exporting Plots in R
You can save your plots using `ggsave()`:

```r
# Save the final plot as a PNG file
ggsave("final_plot.png", plot = final_plot,  width = 15,  height = 10, units = "cm", dpi = 300)
```
